---
title: "Bank Total Assets and Market Cap"
author: "jleonard07"
date: "8/17/2017"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(plotly)
```

## Purpose of Presentation

The purpose of the presentation is to perform an initial review of total assets and market cap for the top 31 banks in the US.

The data set used came from a combination of tables from [List of largest banks in the United States](https://en.wikipedia.org/wiki/List_of_largest_banks_in_the_United_States) on Wikipedia.

## Data process

To prepare the data for plotting these following steps were taken:

1. Join the two tables (from the website) together. This occurred outside of R.
2. Create a factor variable value based on the total assets for a bank.
3. Scale `market_cap` variable.

The code below shows this process:

```{r data, echo=TRUE}
bankData <- read.csv("./bank_assets1.csv", sep = ",", header = TRUE)
bankData$size <- ifelse(bankData$total_assets > 270, "large",
                          ifelse(bankData$total_assets > 100, 
                                 "mid-size", "small"))
bankData$scaled_mc <- sqrt(bankData$market_cap)

```

## Total Assets

The plot below shows the ranks of the banks on the x-axis and the total assets on the y-axis. The size of the bubble indicates its market capital.

```{r ta_plot}
plot_ly(bankData, x = ~rank, y = ~total_assets, type = "scatter",
        mode = 'markers', text = ~name,
        marker = list(size=~scaled_mc, opacity=0.5))
```

## Box Plots on Total Assets
The box plots below are for the three levels of the size factor. The large variance in the *large* group is not surprising - only a handful of banks exceed $1,700b in total assets.

```{r box_ta}
plot_ly(bankData, y = ~total_assets, color = ~size, type = "box")
```

## Box Plots on Market Cap

These show that market cap distributions _mostly_ follow that of total assets. However, there are a couple of outliers.  

```{r box_mc}
plot_ly(bankData, y = ~market_cap, color = ~size, type = "box")
```