---
title: "Imputation and Its Impact on Model Estimation"
author: "jleonard07"
date: "8/26/2017"
output: 
  ioslides_presentation:
    css: styles.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(plotly)
```

## Purpose of Shiny App

The purpose of the Shiny application is to provide a tool to develop intution about how imputation methods can impact model estimation. This will be shown by:

* Allowing the user to select:
      + What percentage of values in a given variable should be replaced with NA.
      + Which statistic, derived from the remaining values of the variable, to use to replace the NAs
* Providing the user the ability to toggle between:
      + Regression results for the estimations using the original data and imputed data
      + Scatter plots of the original data and imputed data


Before proceeding further it should be noted that while the basic idea behind [_imputation_](https://en.wikipedia.org/wiki/Imputation_%28statistics%29) and  [_interpolation_](https://en.wikipedia.org/wiki/Interpolation) is the same (i.e., creating a value where one did not exist before), there is one key distinction. The former **replaces** a missing value with a substitute; the latter **creates** a value between two end points.  

## Imputation Methods

There are a variety of methods used for _single_ imputation (in which only _one_ with missing values gets replacements). Some examples are:


+------------------+------------------------------------+
| Single Imputation| Description                        |
|      Type        |                                    |
+==================+====================================+
| **Mean-value**   | Use the **mean** of non-missing values |
|                  | as replacement                     |
+------------------+------------------------------------+
|**Hot-decking**   | Replace missing values from similar|
|                  | observations _in sample_:          |    
|                  |                                    |   
|                  | + There are different way to do    |
|                  | this                               |
|                  | + Replacing with _out of sample_ is|
|                  | called **cold-decking**            |
+------------------+------------------------------------+
|**Regression**    | Use a univariate or multivariate   |
|                  | regression model to estimate the   |    
|                  | missing variable.                  |
|                  |                                    |
|                  | + The independent variables are 1  |
|                  | or more of the variables in data   |
|                  | set                                |   
+------------------+------------------------------------+

For the purpose of this exploration, **mean-value** and variations on that theme (e.g., min, max, mode) will be used.

## Data Used

To build the intuition the data set used was **LifeCycleSavings** from the R Datasets Package. The data set contains the following variables:

```{r data, echo=TRUE}
str(LifeCycleSavings)
```

For the purpose of exploration, the independent variable will be _ddpi_ (percent growth of disposable income) and the dependent variable will be _sr_ (savings rate).

## Relationship of DDPI to Savings Rate
From the plot below one can see that there is a positive, linear relationship between _ddpi_ and _sr_. The question is: what happens to the relationship when missing values are replaced with the mean, min, max, or median...?

```{r scatter_plot, message=FALSE, fig.height=4, fig.width=7}
p <- plot_ly(data = LifeCycleSavings, x = ~ddpi, y = ~sr)
p
```
